\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}

### 2012-03-29 pkg/evtree/Submission_JSS/Revisions/response_revision_14032012.tex

\documentclass[DIN, pagenumber=false, parskip=half,% 
               fromalign=left, fromphone=true,%  
               fromemail=true, fromurl=false, %
               fromlogo=true, fromrule=false, fromrule=afteraddress]{scrlttr2}

\usepackage[latin1]{inputenc}

\usepackage{graphicx}
\usepackage{amsmath}

\setkomavar{fromname}{Thomas Grubinger\\ Department of Medical Statistics, Informatics and Health Economics \\ Innsbruck Medical University}
\setkomavar{fromaddress}{Sch\"opfstrasse 41/1, 6020 Innsbruck, Austria}
\setkomavar{backaddress}{Dept. of Med. Statistics \\ Innsbruck Medical University\\ Sch\"opfstr. 41/1, A-6020 Innsbruck}
\setkomavar{fromphone}{+43 512 9003 70906}
\setkomavar{fromemail}{thomas.grubinger@i-med.ac.at}
\setkomavar{signature}{Thomas Grubinger, Achim Zeileis und Karl-Peter Pfeiffer}
\setkomavar{fromlogo}{ \includegraphics*[width=2.8cm]{logo}}
\setkomavar{subject}[]{Author's response to the review of ``evtree: 
Evolutionary Learning of Globally Optimal Classification and Regression Trees in R''}

\begin{document}

\begin{letter}{The Editor \\ Journal of Statistical Software}

\opening{Dear Editor}

we would like to thank the reviewer for the careful review and the constructive comments on our paper. We have revised our manuscript based on the reviewer's comments and respond on them below. 
\\ \\
\textbf{1. Overall Thoughts}

\textit{The paper did a good job of introducing a new tree-based algorithm overall. A lot of details pertaining to traditional tree models and the concept of evolutionary algorithms were covered, some of which could be clarified further. A clean-up in notation and wording would be greatly beneficial. The code appears easy-to-read and uses parameters consistent with other tree packages.}

We cleaned-up the notation and wording according to the reviewer's recommendations.
\\ \\

\textit{In the end, the authors gave honest comparisons of their methods to the traditional rpart and ctree methods, but it is unclear as to when and why evtree outperforms its competitors apart from the specific chessboard example given in Section 5.2.}

We added the following paragraph to the ``Conclusions'', as an explanation as to when and why evtree outperforms its competitors:

No obvious relationship between the data characteristics and evtree's 
relative performance could be observed. However, one reason is that for some datasets,
the recursive partitioning trees are already (nearly) optimal. For other datasets, 
 evtree's global search strategy can find more complex interactions, 
resulting in better predictive accuracy and/or sparser trees.
\\ \\
\textbf{2 Spelling/Grammatical Errors}

\textit{
\begin{enumerate}
\item (p.3) Above equation (2), ``predicition" should be ``prediction".
\item (p.7) First paragraph, midway, ``spit points" should be ``split points"
\item (p.8) Minor split rule mutation paragraph: ``less then" should be ``less than", and ``the split
point is change to to..." should be ``the split point is changed to".
\item (p.8) Crossover paragraph, last sentence: ``nods" should be ``nodes".
\item (p.8) Evaluation function first paragraph: should be ``The evaluation function represents the
requirements to which the population should adapt."
\item (p.14) Middle of first pararaph: ``bookclub" should be ``book club".
\end{enumerate}
}

We changed all spelling- and grammatical errors as suggested.
\\  \\
\textbf{3 Notation}

\textit{In Section 2, the authors introduce some notation for the tree setup, but the notation seems inconsistent in the paper. The authors write $\theta=(v_{n_1}, s_{n_1},\dots,v_{n_{M-1}},s_{n_{M-1}})$, where $n_r \in \{1,\dots, M_{max}-1\}$ but then also write $v_r$ and $s_r$ where $r=1,\dots,M-1$. Further, in Section 2.2, the authors talk about $s_r$ and $v_r$ where $r \in \{n_1, \dots, n_{M-1} \}$, which is immediately different from the indexing given
in equation (3). The authors should use a consistent notation for indexing the $v$ and $s$.}

We cleaned up the notation and made it consistent. A binary tree is now denoted as $\theta=(v_1, s_1, \dots, v_{M-1}, s_{M-1})$. Here, $r \in \{1,...,M_{\max}-1\}$ are the internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting \emph{variables}, and $s_r$ the associated \emph{split} rules ($r = 1, \dots, M-1$). Now, in order to simplify the notation, the positions of the internal nodes are no longer made explicit. 
\\  \\
\textbf{4 Case Study}

\textit{
The authors have provided very minimal information on the dataset. (p.14) The authors write
``Thus, the evolutionary tree uses a different cutoff in art for book club members that joined in the last year as opposed to older customers." The text does not explain which variable corresponds to ``join date", but based on Figure 3, one can indirectly infer that this variable is first. The authors have described te variable first as the number of months since the first purchase, but do not say that members join the book club upon making their first purchase.}

We added a Table, which shortly describes the variables of the Bookbinder's Book Club dataset. We assumed the join date to be equal to the data of the first purchase. To avoid confusions we changed the sentence:

``Thus, the evolutionary tree uses a different cutoff in art for book club members that joined in the last year as opposed to older customers."

to:

``Thus, the evolutionary tree uses a different cutoff in art for book club members that bought there first book in the last year as opposed to older customers.".
\\ \\

\textit{The table comparing the misclassification and evaluation function values for the different methods identify the methods as evtree, rpart, ctree, rpart2, and ctree2. The paragraph which describes this table instead identifies the methods as ev, rp, ct, rp2, and ct2. It would be better to make sure the two are consistent.}

We changed the identifiers in the describing paragraph to evtree, rpart, ctree, rpart2, and ctree2. The notation is now consistent.
\\  \\
\textbf{5 Miscellaneous}

In addition to the above, I have the following suggestions.

\begin{enumerate}

\item \textit{(p.7) Section 3.1, ``the assignment of one category is flipped to the other terminal node" might be adjusted to ``one of the c categories is allocated to the other terminal node, to have the effect of ensuring both terminal nodes are nonempty".}

We changed the sentence according to the recommendations.

\textit{\item (p.7) Split operator paragraphs use r again in 2 different ways: ``internal node r" vs. ``internal node at position r". As with the remainder of the paper, r is the internal node, not the position.}

We changed the notation accordingly.

\textit{\item (p.7) Prune paragraph: We might append to the end of the sentence for clarity: ``...and no pruning occurs."}

We changed the sentence according to the recommendations.

\textit{\item (p.8) Minor split rule mutation paragraph: The description of the mutation rules is rather complex, with really 4 different cases to consider (nominal/$<$ 20 values, nominal/$\ge$ 20 values, continuous/$<$ 20 values, continuous/$\ge$ 20 values). It might be good to present this set of rules in a tabular format.}

The rules are now presented in tabular format and are now easier to read.

\textit{\item (p.8) Crossover paragraph, last sentence: ``these are omitted" might be better written as ``they are omitted" to be clear that ``these" refers to the invalid nodes.}

We changed the sentence according to the recommendations.

\textit{\item Evaluation function paragraph: ``A suitable evaluation function... minimizes the models' accuracy...and the models' complexity." It appears that ``minimizes" should be ``optimizes".}

We change the sentence to: ``A suitable evaluation function for classification and regression trees maximizes the models' accuracy on the training data, and minimizes the models' complexity."

\textit{\item (p.8) Classification paragraph: ``misclassification MC" should be ``misclassification rate MC". }

We changed the sentence according to the recommendations.

\textit{\item (p.9) Survivor selection paragraph, last sentence on the page: This sentence is worded awkwardly. Perhaps the authors meant ``In the case of a mutation operator, either the solution before modification, $\theta_i$, or after modification, $\theta_{i+1}$, is kept in memory." Without the commas, the sentence is unclear.}

We changed the sentence according to the recommendations.

\textit{\item (p.10) Third paragraph: Ambiguous ``they" in ``However, as they represent one of the best..., they are both...". Does ``they" refer to the parent and offspring trees? Does this mean that despite the $(1+1)$ rule, both the parent and offspring are both selected to the next generation? This seems unclear.}

Yes, ``they'' refer to the parent and the offspring. However, the sentence explains the strong selective pressure of a $(\mu + \lambda)$ strategy with $\mu>1$ and $\lambda>1$. With a $(1+1)$ strategy, the parent and the offspring can never be selected for the next round (as it is stated in the last sentence of the paragraph).

For better clarification, we replaced ``they'' with ``the parent and the offspring" and changed the sentence:

"Using a $(\mu + \lambda)$ strategy, a strong selective pressure can occur in situations as follows."

with:

"Using a $(\mu + \lambda)$ strategy with $\mu>1$ and $\lambda>1$, a strong selective pressure can occur in situations as follows.".

\textit{\item (p.17) First paragraph: The authors talked about the demanding memory requirements, for which ``400 Mbit" seems very low. Perhaps the authors meant ``400 megabytes"?}

It should be ``400 MB''. We used the wrong abbreviation.
\end{enumerate}
\vspace{0.5cm}

Sincerely,

\vspace{0.75cm}
Thomas Grubinger, Achim Zeileis und Karl-Peter Pfeiffer

\end{letter}

\end{document}

### 2012-03-29 pkg/evtree/inst/doc/evtree.Rnw
almost finished the first revision of the paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@285 255f4249-2263-4417-b1ea-8498779e27a7

In the following, a binary tree model with $M$ terminal nodes (which consequently has
$M-1$ internal splits) is denoted by
%
\begin{equation}
  \theta ~=~ (v_{n_1}, s_{n_1},...,v_{n_{M-1}},s_{n_{M-1}} ), \label{eq:tree}
\end{equation}
%
where the $n_r \in \{1,...,M_{\max}-1\}$ are the positions of the
internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting
\emph{variables}, and $s_r$ the associated \emph{split} rules ($r = 1, \dots, M-1$).
Depending on the domain of $X_{v_r}$, the split rule $s_r$ contains either a cutoff
(for ordered and numeric variables) or a nonempty subset of $\{1, \dots, c\}$
(for a categorical variable with $c$~levels), determining which observations are
sent to the first or second subsample. In the former case, there are $u-1$ possible
split rules if $X_{v_r}$ takes $u$ distinct values; and in the latter case, there
are $2^{c-1} - 1$ possible splits. Thus, the product of all of these combinations
forms all potential elements $\theta$ from $\Theta_M$, the space of conceivable
trees with $M$ terminal nodes. The overall parameter space is then 
$\Theta =\bigcup_{M=1}^{M_{\max}} \Theta_{M}$ (which in practice is often reduced
by exluding elements $\theta$ resulting in too small subsamples etc.).

@@

As done by \cite{breiman1984cart}, let the complexity of a tree be measured by a
function of the number of terminal nodes, without further considering the depth
or the shape of trees. The goal is then to find that classification and regression tree
which optimizes some tradeoff between predicition performance and complexity:
%
\begin{equation} \label{eq:globally}
  \hat{\theta} ~=~ \argmin_{\theta \in \Theta} \; \mathrm{loss}\{Y, f(X,\theta)\} ~+~ \mathrm{comp}(\theta).
\end{equation}
%
where $\mathrm{loss}(\cdot, \cdot)$ is a suitable loss function for the
domain of $Y$; typically, the misclassification rate MC 
and the mean squared error MSE are employed for classification and regression,
respectively. The function $\mathrm{comp}(\cdot)$ is a function that is monotonically
non-decreasing in the number of terminal nodes $M$ of the tree $\theta$, thus penalizing
more complex models in the tree selection selection process.
Note that finding $\hat{\theta}$ requires a search over all $\Theta_M$.

@@

Instead of searching all combinations in $\Theta$ simultaneously, recursive
partitioning algorithms only consider one split at a time. At each internal node
$r \in \{n_1,..., n_{M-1}\}$, the split variable $v_r$ and the corresponding
split point $s_r$ are selected to locally minimize the loss function.
Starting with an empty tree $\theta_0 = (\emptyset)$, the tree is first grown 
recursively and subsequently \emph{pruned} to satisfy the complexity tradeoff:
%
\begin{eqnarray} \label{eq:locally}
  \tilde \theta_r
    & = & \argmin_{ \theta = \theta_{r-1} \cup (v_r, s_r)} \; \mathrm{loss}\{Y, f(X, \theta)\} 
    \qquad (r = 1, \dots, M_{\max} -1), \\
  \tilde \theta
    & = & \argmin_{\tilde \theta_r} \; \mathrm{loss}\{Y, f(X, \tilde \theta_r)\} ~+~ \mathrm{comp}(\tilde \theta_r).
  \label{eq:pruning}
\end{eqnarray}
%
For nontrivial problems, forward-search recursive partitioning methods only search a small
fraction of the global search space $(v_1, s_1, \dots, v_{M_{\max}-1},
s_{M_{\max}-1} )$. They only search each $(v_r, s_r)$ once, and independently of
the subsequent split rules, hence typically leading to a globally suboptimal solution
$\tilde \theta$.

@@

\setkeys{Gin}{width=0.5\textwidth}
\begin{figure}[t!]
\centering
<<chess22-plot, fig=TRUE, echo=FALSE, height=4, width=4>>=
plot(X2 ~ X1, data = chess22, xlim = c(0, 2), ylim = c(0, 2), pch = c(1, 4)[Y], col = c("black", "slategray")[Y])
@
\caption{\label{fig:illustration} Class distribution of  the $(X_1, X_2)$-plane. The two classes are indicated by black circles and gray crosses.}
\vspace*{-0.3cm} %%FIXME: to avoid \pagebreak
\end{figure}

@@

Each tree of the population is initialized with a valid, randomly generated,
split rule in the root node. First, $v_1$ is selected with uniform probability
from $1,...,P$. Second, a split point $s_1$ is selected. If $X_{v_1}$ is numeric
or ordinal with $u$ unique values, a split point $s_1$ is selected with uniform
probability from the $u-1$ possible spit points of $X_{v_1}$. If $X_{v_1}$ is
nominal and has $c$ categories, each $k=1,...,c$ has a probability of $50\%$ to
be assigned to the left or the right daughter node. In cases where all $k$ are
allocated to the same terminal node, the assignment of one category is flipped
to the other terminal node. If this procedure results in a non-valid split rule,
the two steps of random split variable selection and split point selection are
repeated. With the definition of $r=1$ and the selection of $v_1$ and $s_1$, the
initialization is complete and each individual of the population of trees is of
type ${\theta}_{1}= (v_1, s_1)$.

@@

The search for a valid split rule is conducted as in Section~\ref{init}
for a maximum of $P$~iterations. 
In cases where no valid split rule can be assigned to the internal node at
position~$r$, the search for a valid split rule is carried out on another
randomly selected terminal node. If, after 10~attempts, no valid split rule
can be found, then ${\theta}_{i+1} = {\theta}_{i}$. Otherwise, the
set of parameters in iteration $i+1$ are given by ${\theta}_{i+1}=
{\theta}_i \cup (v_{r},s_{r})$.

@@

\textit{Prune} chooses a random internal node $r$, where $r>1$, which has two
terminal nodes as successors and prunes it into a terminal node. The tree's
parameters at iteration $i+1$ are ${\theta}_{i+1}={\theta}_{i}
\setminus (v_{r},s_{r})$. If ${\theta}_{i}$ only comprises one internal
node, i.e., the root node, then ${\theta}_{i+1} = {\theta}_{i}$.

@@

\textit{Minor split rule mutation} is similar to the \textit{major split rule
mutation} operator, but it does not alter $v_r$ and only changes the split point
$s_r$ by a minor degree. If $X_{v_r}$ is numerical or ordinal the split point
$s_r$ is changed by a non-zero number of unique values of $X_{v_r}$. In cases
where $X_{v_r}$ has less then $20$ unique values, the split point is changed to
the next larger, or the next lower, unique value of $X_{v_r}$. Otherwise,
$s_r$ is randomly shifted by a number of unique values that is not larger than
$10\%$ of the range of unique values of $X_{v_r}$. If $X_{v_r}$ is a nominal
variable, with less than $20$ categories, one of the categories is randomly
modified. Otherwise, at least one and at most $10\%$ of the variable's
categories are changed. In cases where subsequent nodes become invalid, further
split points are searched that preserve the tree's topology. After five
non-successful attempts at finding a topology preserving split point, the
non-valid nodes are pruned.

@@

\textit{Crossover} exchanges, randomly selected, subtrees between two trees. Let
${\theta}_{i}^1$ and ${\theta}_{i}^2$ be the two trees
chosen from the population for crossover. First, two internal nodes $r_1$ and $r_2$
are selected randomly from ${\theta}_{i}^1$ and ${\theta}_{i}^2$, respectively.
Let $\mathrm{sub}1({\theta}_i^j, r_j)$ denote the subtree of $\theta_i^j$ rooted
by $r_j$ ($j=1,2$), i.e., the tree containing $r_j$ and its descendant nodes.
Then, the complementary part of ${\theta}_{i}^j$ can be defined as 
$\mathrm{sub}2({\theta}_{i}^j, r_j) = {\theta}_{i}^j \setminus \mathrm{sub}1({\theta}_{i}^j, r_j)$. 
The crossover operator creates two new trees
${\theta}_{i+1}^1 = \mathrm{sub}2({\theta}_{i}^1, r_1) \cup \mathrm{sub}1({\theta}_{i}^2, r_2)$ and
${\theta}_{i+1}^2 = \mathrm{sub}2({\theta}_{i}^2, r_2) \cup \mathrm{sub}1({\theta}_{i}^1, r_1)$.
If the crossover creates some invalid nods in either one of the new trees ${\theta}_{i+1}^1$
or ${\theta}_{i+1}^2$, these are omitted.

@@

The evaluation function represents the requirements the population should adapt
to. In general, these requirements are formulated by Equation~\ref{eq:globally}.
A suitable evaluation function for classification and regression trees minimizes
the models' accuracy on the training data, and the models' complexity. This
subsection describes the currently implemented choices of evaluation functions
for classification and for regression.

@@

The quality of a classification tree is most commonly measured as a function of
its misclassification MC and the complexity of a tree by a function
of the number of its terminal nodes $M$. \code{evtree} uses $2 N \cdot
\mathrm{MC}(Y, f(X, \theta))$ as a loss function. The number of terminal nodes,
weighted by $\log N$ and a user-specified parameter $\alpha$, measures the
complexity of trees.
%
\begin{eqnarray}
  \mathrm{loss}(Y, f(X, \theta)) & = & 2 N \cdot \mathrm{MC}(Y, f(X, \theta)) \nonumber \\
                                 & = & 2 \cdot \sum_{n=1}^N I(Y_n \neq f(X_{\cdot n}, \theta)),  \label{eq:classification} \\
  \mathrm{comp}(\theta)          & = & \alpha \cdot M \cdot \log N. \nonumber
\end{eqnarray}
%
With these particular choices, Equation~\ref{eq:globally} seeks trees
$\hat{\theta}$ that minimize the misclassification loss at a BIC-type tradeoff
with the number of terminal nodes.

@@

Our algorithm uses $(1+1)$ selection, where one parent solution
competes with one offspring for a place in the population. In the case of a
mutation operator either the solution before ${\theta}_i$ or after
modification ${\theta}_{i+1}$ is kept in memory. In the case of the
crossover operator, the initial solutions of ${\theta}_i^1$ competes with
its subsequent solutions ${\theta}_{i+1}^1$. Correspondingly, one of the
two solutions ${\theta}_i^2$ and ${\theta}_{i+1}^2$ is rejected. The
survivor selection is done deterministically. The tree with lower fitness,
according to the evaluation function, is rejected. Note that, due to the
definition of the crossover operator, some trees are selected more than
once in an iteration. Correspondingly, these trees undergo the survival
selection process more than once in an iteration.

@@

There are two important issues in the evolution process of an evolutionary
algorithm: population diversity and selective pressure
\citep{michalewicz1994genetic}. These factors are related, as with increasing
selective pressure the search is focused more around the currently best
solutions. An overly strong selective pressure can cause the algorithm to
converge early in local optima. On the other hand, an overly weak selective
pressure can make the search ineffective. Using a $(\mu + \lambda)$ strategy, a
strong selective pressure can occur in situations as follows. Suppose the $b$-th
tree of the population is one of the fittest trees in iteration $i$, and in
iteration $i$  one split rule of the $b$-th tree is changed only by a minor
degree. Then very few instances are classified differently and the overall
misclassification might not even change. However, as they represent one of the
best solutions in iteration $i$, they are both selected for the subsequent
population. This situation can occur frequently, especially when a fine-tuning
operator like the \textit{minor split rule mutation} is used. Then, the
diversity of different trees is lost quickly and the algorithm likely terminates
in a local optimum. The $(1+1)$ selection mechanism clearly avoids these
situations, as only the parent or the offspring can be part of the subsequent
population.

@@

An interesting application for classification tree analysis is target marketing,
where limited resources are aimed at a distinct group of potential customers. An
example is provided by \cite{lilien2004} in the \textit{Bookbinder's Book Club}
marketing case study about a (fictitious) American book club. In this case study, a brochure of the book
``The Art History of Florence'' was sent to 20,000~customers, 1,806~of
which bought the book. The dataset contains a subsample of 1,300~customers for
building a predictive model of customer choice.

@@

Disadvantages of the \code{evtree} algorithm are computation time and memory requirements. While the smallest of the analyzed datasets, \textit{Glass identification}, only needed approximately 4--6~seconds to fit, larger datasets demanded several minutes. The fit of a model from the largest dataset, \textit{MAGIC gamma telescope}, required approximately 40--50 minutes and a main memory of 400 Mbit. The required resources were measured on an Intel Core 2 Duo with 2.2 GHz and 2 GB RAM using the 64-bit version of Ubuntu 10.10.

@@

@@

>>>>

In the following, a binary tree model with $M$ terminal nodes (which consequently has
$M-1$ internal splits) is denoted by
%
\begin{equation}
  \theta ~=~ (v_{1}, s_{1},...,v_{M-1},s_{M-1} ), \label{eq:tree}
\end{equation}
%
where $r \in \{1,...,M_{\max}-1\}$ are the internal \emph{nodes}, $v_r \in \{1, \dots ,P\}$ the associated splitting
\emph{variables}, and $s_r$ the associated \emph{split} rules ($r = 1, \dots, M-1$).
Depending on the domain of $X_{v_r}$, the split rule $s_r$ contains either a cutoff
(for ordered and numeric variables) or a nonempty subset of $\{1, \dots, c\}$
(for a categorical variable with $c$~levels), determining which observations are
sent to the first or second subsample. In the former case, there are $u-1$ possible
split rules if $X_{v_r}$ takes $u$ distinct values; and in the latter case, there
are $2^{c-1} - 1$ possible splits. Thus, the product of all of these combinations
forms all potential elements $\theta$ from $\Theta_M$, the space of conceivable
trees with $M$ terminal nodes. The overall parameter space is then 
$\Theta =\bigcup_{M=1}^{M_{\max}} \Theta_{M}$ (which in practice is often reduced
by exluding elements $\theta$ resulting in too small subsamples etc.).

@@

As done by \cite{breiman1984cart}, let the complexity of a tree be measured by a
function of the number of terminal nodes, without further considering the depth
or the shape of trees. The goal is then to find that classification and regression tree
which optimizes some tradeoff between prediction performance and complexity:
%
\begin{equation} \label{eq:globally}
  \hat{\theta} ~=~ \argmin_{\theta \in \Theta} \; \mathrm{loss}\{Y, f(X,\theta)\} ~+~ \mathrm{comp}(\theta).
\end{equation}
%
where $\mathrm{loss}(\cdot, \cdot)$ is a suitable loss function for the
domain of $Y$; typically, the misclassification rate MC 
and the mean squared error MSE are employed for classification and regression,
respectively. The function $\mathrm{comp}(\cdot)$ is a function that is monotonically
non-decreasing in the number of terminal nodes $M$ of the tree $\theta$, thus penalizing
more complex models in the tree selection selection process.
Note that finding $\hat{\theta}$ requires a search over all $\Theta_M$.

@@

Instead of searching all combinations in $\Theta$ simultaneously, recursive
partitioning algorithms only consider one split at a time. At each internal node
$r \in \{1,...,M-1\}$, the split variable $v_r$ and the corresponding
split point $s_r$ are selected to locally minimize the loss function.
Starting with an empty tree $\theta_0 = (\emptyset)$, the tree is first grown 
recursively and subsequently \emph{pruned} to satisfy the complexity tradeoff:
%
\begin{eqnarray} \label{eq:locally}
  \tilde \theta_r
    & = & \argmin_{ \theta = \theta_{r-1} \cup (v_r, s_r)} \; \mathrm{loss}\{Y, f(X, \theta)\} 
    \qquad (r = 1, \dots, M_{\max} -1), \\
  \tilde \theta
    & = & \argmin_{\tilde \theta_r} \; \mathrm{loss}\{Y, f(X, \tilde \theta_r)\} ~+~ \mathrm{comp}(\tilde \theta_r).
  \label{eq:pruning}
\end{eqnarray}
%
For nontrivial problems, forward-search recursive partitioning methods only search a small
fraction of the global search space $(v_1, s_1, \dots, v_{M_{\max}-1},
s_{M_{\max}-1} )$. They only search each $(v_r, s_r)$ once, and independently of
the subsequent split rules, hence typically leading to a globally suboptimal solution
$\tilde \theta$.

@@

\setkeys{Gin}{width=0.5\textwidth}
\begin{figure}[t!]
\centering
<<chess22-plot, fig=TRUE, echo=FALSE, height=4, width=4>>=
plot(X2 ~ X1, data = chess22, xlim = c(0, 2), ylim = c(0, 2), pch = c(1, 4)[Y], col = c("black", "slategray")[Y])
@
\caption{\label{fig:illustration} Class distribution of  the $(X_1, X_2)$-plane. The two classes are indicated by black circles and gray crosses.}
%%\vspace*{-0.3cm} %%FIXME: to avoid \pagebreak
\end{figure}

@@

Each tree of the population is initialized with a valid, randomly generated,
split rule in the root node. First, $v_1$ is selected with uniform probability
from $1,...,P$. Second, a split point $s_1$ is selected. If $X_{v_1}$ is numeric
or ordinal with $u$ unique values, a split point $s_1$ is selected with uniform
probability from the $u-1$ possible split points of $X_{v_1}$. If $X_{v_1}$ is
nominal and has $c$ categories, each $k=1,...,c$ has a probability of $50\%$ to
be assigned to the left or the right daughter node. In cases where all $k$ are
allocated to the same terminal node, one of the $c$ categories is allocated to 
the other terminal node, to have the effect of ensuring both terminal nodes are nonempty.
If this procedure results in a non-valid split rule, the two steps of random split variable 
selection and split point selection are repeated. With the definition of $r=1$ and the 
selection of $v_1$ and $s_1$, the initialization is complete and each individual of
the population of trees is of type ${\theta}_{1}= (v_1, s_1)$.

@@

The search for a valid split rule is conducted as in Section~\ref{init}
for a maximum of $P$~iterations. 
In cases where no valid split rule can be assigned to internal node~$r$,
the search for a valid split rule is carried out on another
randomly selected terminal node. If, after 10~attempts, no valid split rule
can be found, then ${\theta}_{i+1} = {\theta}_{i}$. Otherwise, the
set of parameters in iteration $i+1$ are given by ${\theta}_{i+1}=
{\theta}_i \cup (v_{r},s_{r})$.

@@

\textit{Prune} chooses a random internal node $r$, where $r > 1$, which has two
terminal nodes as successors and prunes it into a terminal node. The tree's
parameters at iteration $i+1$ are ${\theta}_{i+1}={\theta}_{i}
\setminus (v_{r},s_{r})$. If ${\theta}_{i}$ only comprises one internal
node, i.e., the root node, then ${\theta}_{i+1} = {\theta}_{i}$ and no pruning occurs.

@@

\textit{Minor split rule mutation} is similar to the \textit{major split rule
mutation} operator. However, it does not alter $v_r$ and only changes the split point
$s_r$ by a minor degree, which is defined by one of the following 4 cases:
\begin{itemize}
\item {$X_{v_r}$ is numerical or ordinal and has at least $20$ unique values:}
The split point $s_r$ is randomly shifted by a non-zero number of unique values of $X_{v_r}$ 
that is not larger than $10\%$ of the range of unique values.
\item {$X_{v_r}$ is numerical or ordinal and has less than $20$ unique values:}
The split point is changed to the next larger, or the next lower, unique value of $X_{v_r}$.
\item {$X_{v_r}$ is nominal and has at least $20$ categories:}  at least one and at most $10\%$ of the variable's
categories are changed.
\item {$X_{v_r}$ is nominal and has less than $20$ categories:} One of the categories is randomly
modified.
\end{itemize}
In cases where subsequent nodes become invalid, further
split points are searched that preserve the tree's topology. After five
non-successful attempts at finding a topology preserving split point, the
non-valid nodes are pruned.

@@

\textit{Crossover} exchanges, randomly selected, subtrees between two trees. Let
${\theta}_{i}^1$ and ${\theta}_{i}^2$ be the two trees
chosen from the population for crossover. First, two internal nodes $r_1$ and $r_2$
are selected randomly from ${\theta}_{i}^1$ and ${\theta}_{i}^2$, respectively.
Let $\mathrm{sub}1({\theta}_i^j, r_j)$ denote the subtree of $\theta_i^j$ rooted
by $r_j$ ($j=1,2$), i.e., the tree containing $r_j$ and its descendant nodes.
Then, the complementary part of ${\theta}_{i}^j$ can be defined as 
$\mathrm{sub}2({\theta}_{i}^j, r_j) = {\theta}_{i}^j \setminus \mathrm{sub}1({\theta}_{i}^j, r_j)$. 
The crossover operator creates two new trees
${\theta}_{i+1}^1 = \mathrm{sub}2({\theta}_{i}^1, r_1) \cup \mathrm{sub}1({\theta}_{i}^2, r_2)$ and
${\theta}_{i+1}^2 = \mathrm{sub}2({\theta}_{i}^2, r_2) \cup \mathrm{sub}1({\theta}_{i}^1, r_1)$.
If the crossover creates some invalid nodes in either one of the new trees ${\theta}_{i+1}^1$
or ${\theta}_{i+1}^2$, they are omitted.

@@

The evaluation function represents the requirements to which the population should adapt. 
In general, these requirements are formulated by Equation~\ref{eq:globally}.
A suitable evaluation function for classification and regression trees maximizes
the models' accuracy on the training data, and minimizes the models' complexity. This
subsection describes the currently implemented choices of evaluation functions
for classification and for regression.

@@

The quality of a classification tree is most commonly measured as a function of
its misclassification rate MC and the complexity of a tree by a function
of the number of its terminal nodes $M$. \code{evtree} uses $2 N \cdot
\mathrm{MC}(Y, f(X, \theta))$ as a loss function. The number of terminal nodes,
weighted by $\log N$ and a user-specified parameter $\alpha$, measures the
complexity of trees.
%
\begin{eqnarray}
  \mathrm{loss}(Y, f(X, \theta)) & = & 2 N \cdot \mathrm{MC}(Y, f(X, \theta)) \nonumber \\
                                 & = & 2 \cdot \sum_{n=1}^N I(Y_n \neq f(X_{\cdot n}, \theta)),  \label{eq:classification} \\
  \mathrm{comp}(\theta)          & = & \alpha \cdot M \cdot \log N. \nonumber
\end{eqnarray}
%
With these particular choices, Equation~\ref{eq:globally} seeks trees
$\hat{\theta}$ that minimize the misclassification loss at a BIC-type tradeoff
with the number of terminal nodes.

@@

Our algorithm uses $(1+1)$ selection, where one parent solution
competes with one offspring for a place in the population. In the case of a
mutation operator, either the solution before modification, ${\theta}_i$, or after
modification, ${\theta}_{i+1}$, is kept in memory. In the case of the
crossover operator, the initial solutions of ${\theta}_i^1$ competes with
its subsequent solutions ${\theta}_{i+1}^1$. Correspondingly, one of the
two solutions ${\theta}_i^2$ and ${\theta}_{i+1}^2$ is rejected. The
survivor selection is done deterministically. The tree with lower fitness,
according to the evaluation function, is rejected. Note that, due to the
definition of the crossover operator, some trees are selected more than
once in an iteration. Correspondingly, these trees undergo the survival
selection process more than once in an iteration.

@@

There are two important issues in the evolution process of an evolutionary
algorithm: population diversity and selective pressure
\citep{michalewicz1994genetic}. These factors are related, as with increasing
selective pressure the search is focused more around the currently best
solutions. An overly strong selective pressure can cause the algorithm to
converge early in local optima. On the other hand, an overly weak selective
pressure can make the search ineffective. Using a $(\mu + \lambda)$ strategy 
with $\mu>1$ and $\lambda>1$, a strong selective pressure can occur in 
situations as follows. Suppose the $b$-th
tree of the population is one of the fittest trees in iteration $i$, and in
iteration $i$ one split rule of the $b$-th tree is changed only by a minor
degree. Then very few instances are classified differently and the overall
misclassification might not even change. However, as the parent and the offspring
represent one of the best solutions in iteration $i$, they are both selected for the subsequent
population. This situation can occur frequently, especially when a fine-tuning
operator like the \textit{minor split rule mutation} is used. Then, the
diversity of different trees is lost quickly and the algorithm likely terminates
in a local optimum. The $(1+1)$ selection mechanism clearly avoids these
situations, as only the parent or the offspring can be part of the subsequent
population.

@@

An interesting application for classification tree analysis is target marketing,
where limited resources are aimed at a distinct group of potential customers. An
example is provided by \cite{lilien2004} in the \textit{Bookbinder's Book Club}
marketing case study about a (fictitious) American book club. In this case study, a brochure of the book
``The Art History of Florence'' was sent to 20,000~customers, 1,806~of
which bought the book. The dataset contains a subsample of 1,300~customers with 10~exploratory variables (see Table~\ref{tab:BBBClubDescription}) for building a predictive model of customer choice.

\begin{table}[t!]
\centering
\begin{tabular}{ll}
\toprule	
     \textbf{Variable} & Description\\
\midrule	
     \textbf{amount} & total amount of money spent at the BBB Club. \\
     \textbf{art} & number of art books purchased. \\
     \textbf{child} & number of children's books purchased. \\
     \textbf{cook} & number of cookbooks purchased. \\
     \textbf{diy} & number of do-it-yourself books purchased. \\    
     \textbf{first} & number of months since the first purchase. \\   
     \textbf{freq} & number of books purchased at the BBB Club. \\          
     \textbf{gender} & factor indicating gender. \\
     \textbf{last} & number of months since the last purchase. \\         
     \textbf{youth} & number of youth books purchased. \\
\bottomrule
\end{tabular}
\caption{\label{tab:BBBClubDescription} This table describes the predictor variables of the \textit{Bookbinder's Book Club} dataset.} 
\end{table}

@@

Disadvantages of the \code{evtree} algorithm are computation time and memory requirements. While the smallest of the analyzed datasets, \textit{Glass identification}, only needed approximately 4--6~seconds to fit, larger datasets demanded several minutes. The fit of a model from the largest dataset, \textit{MAGIC gamma telescope}, required approximately 40--50 minutes and a main memory of 400 MB. The required resources were measured on an Intel Core 2 Duo with 2.2 GHz and 2 GB RAM using the 64-bit version of Ubuntu 10.10.

@@

%%\pagebreak %%FIXME

@@

No obvious relationship between the data characteristics and \code{evtree}'s 
relative performance could be observed. However, one reason is that for some datasets,
the recursive partitioning trees are already (nearly) optimal. For other datasets, 
 \code{evtree}'s global search strategy can find more complex interactions, 
resulting in better predictive accuracy and/or sparser trees.
